---
output: github_document
---

[![lifecycle](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![Travis build status](https://travis-ci.org/cboettig/taxald.svg?branch=master)](https://travis-ci.org/cboettig/taxald)
[![Coverage status](https://codecov.io/gh/cboettig/taxald/branch/master/graph/badge.svg)](https://codecov.io/github/cboettig/taxald?branch=master)
[![CRAN status](https://www.r-pkg.org/badges/version/taxald)](https://cran.r-project.org/package=taxald)


<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
# taxald

**This package is purley exploratory at this stage!**



Central to `taxald` design is to create a *local* database of all avialable taxonomic authority data that we can query quickly, even when working with tens of thousands of species. The second key feature of `taxald` design is to return queries from each authority using a *consistent* and convenient table structure, facilitating queries that compare, combine, or work across names from multiple authorities.  

## Install and initial setup

`taxald` should now be installable.  To get started, try:

```{r eval=FALSE}
devtools::install_github("cboettig/taxald")
```

Before we can use most `taxald` functions, we need to do a one-time installation of the database `taxald` uses for almost all commands.  This can take a while to run, but needs only be done once.  The database is installed on your local harddisk and will persist between R sessions.  Download and install the database 

```{r}
library(taxald)


```


## Package API

Design sketch/spec for package API:

- Given vector of taxonomic names, return taxonomic identifiers
- Given taxonomic identifiers, return heirarchical classification.  

(Note that for some authorities, e.g. GBIF and TPL, taxonomic identifiers are only assigned to species names.  So given the name of a higher-level rank, no id could be returned)

- Given any higher-order rank name (e.g. `infraorder`) return identifiers of all member species.

- Given common names, resolve to accepted scientific name

- Resolve any known miss-spelling, recognized synonymous name, generic name or name-part to corresponding taxonomic name.

- Map between identifiers

- Normalize rank names

- Resolve a list of names across multiple authorities to attain higher coverage, 
- provide merged tables


See [schema.md](schema.md) for a sketch of the underlying database architecture imposed on the data source from each authority. 

Note that all inputs and outputs should be vectorized.  
