---
output: github_document
---

[![lifecycle](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![Travis build status](https://travis-ci.org/cboettig/taxald.svg?branch=master)](https://travis-ci.org/cboettig/taxald)
[![Coverage status](https://codecov.io/gh/cboettig/taxald/branch/master/graph/badge.svg)](https://codecov.io/github/cboettig/taxald?branch=master)
[![CRAN status](https://www.r-pkg.org/badges/version/taxald)](https://cran.r-project.org/package=taxald)


<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
# taxald

**This package is purley exploratory at this stage!**

Central to `taxald` design is to create a *local* database of all available taxonomic authority data that we can query quickly, even when working with tens of thousands of species. The second key feature of `taxald` design is to return queries from each authority using a *consistent* and convenient table structure, facilitating queries that compare, combine, or work across names from multiple authorities.  

## Install and initial setup

`taxald` should now be installable.  To get started, try:

```{r eval=FALSE}
devtools::install_github("cboettig/taxald")
```

```{r}
library(taxald)
library(tictoc) # let's display timing of queries for reference
```

Before we can use most `taxald` functions, we need to do a one-time installation of the database `taxald` uses for almost all commands.  This can take a while to run, but needs only be done once.  The database is installed on your local harddisk and will persist between R sessions.  We download and install the database, specifying this location:

```{r }
td_create(dbdir = "~/.taxald")
```

The default behavior installs only the ITIS database using the hierachy schema.  While this should be sufficient for getting started on basic tasks like looking up classification or translating between IDs and scientific names, users will want to install additional authorities and schemas for advanced use. See the `authorities` vignette for details.  By default, `taxald` will install the database into a hidden `.taxald` folder in your home directory (`~/.taxald`).  You can control this by passing the path to `create_taxadb()` or setting the environmental variable `TAXALD_HOME` to a different location.  

This function downloads each of the individual tables and loads them into a [MonetDBLite](https://www.monetdb.org) database.  (MonetDBLite is much like SQLite, requiring no external server setup.  However, for our purposes, MonetDB's columnar design provides significantly faster performance than SQLite, Postgres, or even some in-memory operations in `dplyr`.)  

Currently tables are pulled from the development cache on GitHub. Many of these tables are still being cleaned up and standardized, see [schema.md](schema.md).  Multiple tablesare provided for each authority, though some of these contain redundant information, different orientations may prove more convenient or computationally efficient.  



## Package API

Once the database is installed, we can start to make some queries. 

First, let's get a nice big species list, say, all the birds (known to ITIS):

```{r}
tic()
df <- descendants(name = "Aves", rank = "class")
toc()
```

How many species did we get?

```{r}
length(df$species)
```

Generally, we have a species list could have come from some particular research project, and the names may not match those of any one particular authority.

```{r}
tic()
ids(df$species)
toc()
```


We can get the full hierachical classification for this list of species:  

```{r}
tic()
classification(df$species)
toc()
```

Design sketch/spec for package API:

- Given vector of taxonomic names, return taxonomic identifiers
- Given taxonomic identifiers, return heirarchical classification.  

(Note that for some authorities, e.g. GBIF and TPL, taxonomic identifiers are only assigned to species names.  So given the name of a higher-level rank, no id could be returned)

- Given any higher-order rank name (e.g. `infraorder`) return identifiers of all member species.

- Given common names, resolve to accepted scientific name

- Resolve any known miss-spelling, recognized synonymous name, generic name or name-part to corresponding taxonomic name.

- Map between identifiers

- Normalize rank names

- Resolve a list of names across multiple authorities to attain higher coverage, 
- provide merged tables


See [schema.md](schema.md) for a sketch of the underlying database architecture imposed on the data source from each authority. 

Note that all inputs and outputs should be vectorized.  
