---
title: "taxald tutorial"
author: "Carl Boettiger"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{taxald}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```



```{r}
library(taxald)
library(dplyr)
library(tidyr)
```



# Introduction: the problem of name matching

We frequently want to combine different data sources using species names.  Perhaps we have one table which gives us occurrance information for a given list of species, and another table that contains trait information for species, and maybe a third source that provides a phylogenetic tree.  If our system of scientific species names was perfect, we could simply do a "table join" using species name as the joining `key`, and all would be well. 

Unfortunately, as anyone who has attempted this kind of exercise over more than a handful of species has discovered, this often works for most but very rarely for all of the species names involved.  There are several reasons this process runs into problems.  

**Synonyms**.  One of the most common problems is the existence of synonyms: different names or different spellings of a given scientific name.  While this could include definite miss-spellings, for there are many species for which authorities can differ in the preferred name -- essentially, two or more different names correspond to the same species, and should be treated as such in our species join.  


For instance, we consider the 233 primate species names used in the `geiger` package `primates` data.  To avoid installing that dependency-heavy package, a copy of these names has been cached in `taxald` and can be loaded as follows: 

```{r}
ex <- system.file("extdata", "primates.tsv.bz2", package = "taxald")
primates <- readr::read_tsv(ex)
primates
```

Our goal will be to associate each name with a definitive taxonomic ID of an existing naming authority.  This will allow us to merge on IDs directly, rather than names.  By mapping both recognized names and synonyms to the same corresponding taxonomic ID, we can be sure that we can join the relevant data correctly.


## Setup

To get started, we'll install all the data sources available to `taxald` into a local database. This may take a while, particularly over a slow internet connection, but it needs to be done only once.  The downloaded size of all data is around 3 GB.  Once this task completes, our subsequent operations should all be quite fast.  

```{r}
td_create(authorities = "all")
```


## Resolving names

**Developer note: the following outlines the current mechanism for name resolution, which queries the recognized names and synonyms (if present) of each authority sequentially.  While this is illustrative of the underlying process, we probably want a helper function which automates this.**

Match a list of 233 species names against a naming authority:  


```{r}
my_taxa <- primates %>%  # 233 taxa
  mutate(id = ids(species)) 

is.na(my_taxa$id) %>% sum()
```
We manage to resolve 230 of the 233 species.  Not bad.  


By default, `taxald::ids()` function automatically resolves names against both ids and any synonyms recognized by the requested authority (ITIS). If we did not resolve synonyms, we find many more of the names would not match: 

```{r}
tmp <- primates %>%  
  mutate(id = ids(species, 
                  synonyms_check = FALSE)) 

is.na(tmp$id) %>% sum()
```




```{r}

my_taxa <- primates %>%  
  mutate(itis = ids(species, "itis"),
         ncbi = ids(species, "ncbi"),
         col = ids(species, "col"),
         gbif = ids(species, "gbif"),
         wd = ids(species, "wd"),
         tpl = ids(species, "tpl"),
         fb = ids(species, "fb"),
         slb = ids(species, "slb")) 

```

Which IDs are missing?

```{r}
my_taxa$has_match <- my_taxa %>% 
  select(-species) %>%
  purrr::map_dfc(function(x) !is.na(x)) %>% 
  rowSums() > 0
```

Looks like wikidata covers the most, including anything covered by at least one other authority: 

```{r}
missing_wd <- my_taxa %>% filter(is.na(wd)) %>% pull(species)
missing_all <- my_taxa %>% filter(!has_match) %>% pull(species)

still_missing <- identical(missing_wd, missing_all)

```

Let's try the synonyms tables from the other authorities: 
