---
title: "Crosswalking The Authorities"
author: "Carl Boettiger"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{crosswalk}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```



```{r message = FALSE}
library(taxald)
library(dplyr)
library(tidyr)
```

Work in progress -- the mismatches of `itis`, `col` and `wd` here illustrate the complexity of name matching.  Some names resolve multiple times.  Some authorities (ITIS) recognize certain synonyms that they still haven't mapped to accepted id.  

```{r}
fish <- taxa_tbl("fb") %>%  
  select(fb = id, species) %>% 
  collect() 
species <- pull(fish, species)

itis <- ids(species, "itis")
wd <- ids(species, "wd")
col <- ids(species, "col")

## Note each of these authorities all return more rows than we had in the input table!
sapply(list(fb = fish, itis = itis, wd = wd, col = col), function(x) length(x[[1]]))

```


```{r}
my_taxa <- fish %>%

  mutate(#itis = get_ids(species, "itis"),
         ncbi = get_ids(species, "ncbi"),
         #col = get_ids(species, "col"),
         gbif = get_ids(species, "gbif"),
         #wd = get_ids(species, "wd"),
         tpl = get_ids(species, "tpl"),
        # fb = get_ids(species, "fb"),
         slb = get_ids(species, "slb")) 
```


```{r}
my_taxa %>% 
  select(-species) %>% 
  purrr::map_dbl(function(x) sum(!is.na(x)))
```

Looks like three plants have matching scientific names to some of our fish:

```{r}
dup <- fish %>% pull(species) %>% ids(authority = "tpl") %>% filter(!is.na(id))
dup
```

This also probably explains why `col` and `wd` are returning the wrong-length matches:

```{r}
species <- pull(fish, species)

col_fb <- ids(species, "col") 
dim(col_fb)[1] - length(species)

wd_fb <- ids(species, "wd") 
dim(wd_fb)[1] - length(species)

```

```{r}

#col_hierarchy <- classification(id = col_fb$id) %>% filter(kingdom == "Animalia") 

taxa_tbl("col", "hierarchy") %>%
  filter(kingdom == "Animalia") %>%
  semi_join(select(col_fb, id), copy = TRUE) %>%
  select(id, species) %>% 
  collect()
```



```{r}
has_match <- my_taxa %>% 
  select(-species, -fb) %>%
  purrr::map_dfc(function(x) !is.na(x)) %>% 
  rowSums() > 0

my_taxa %>% filter(!has_match)
```
