---
title: "Database schema for taxadb"
author: "Carl Boettiger"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{schema}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```



`taxadb` relies on a set of pre-assembled tables following a set of standardized schema layouts using Darwin Core vocabulary, as outlined below.  The database dumps provided by providers supported in `taxadb` at this time are:

- `itis`: [The Integrated Taxonomic Information System](https://www.itis.gov/)
- `col`: [The Catalogue of Life](http://www.catalogueoflife.org/)
- `ncbi`: [The National Center for Biotechnology Information](https://www.ncbi.nlm.nih.gov/)
- `gbif`: [The Global Biodiversity Information Facility](https://www.gbif.org/)
- `tpl`: [The Plant List](http://www.theplantlist.org/)
- `fb`: [FishBase](https://fishbase.ca)
- `slb`: [SeaLifeBase](https://www.sealifebase.ca/)
- `wd`: [WikiData](https://wikidata.org)
- `iucn`: The IUCN Red List of endangered species status
- `ott`: Open Tree of Life taxonomy.  

Note that `col`, `gbif`, `ott`, are explicitly synthesis projects integrating the databases of names from a range of (many) other providers, while `itis`, `iucn`, `ncbi`, `tpl`, `fb`, and `slb` are independent name providers.  The synthetic or integrated name lists are not a complete substitute for the databases they integrate, owing to both differences in how up-to-date the relative records are as well as to either expert disagreements or algorithmic miss-matches.  `taxadb` makes no attempt to provide an opinion or reconciliation mechanism to any of these issues, but only to provide convenient access to data and functions for manipulating these records in a fast and consistent manner.

These providers provide taxonomic data in a wide range of database formats using a wide range of data layouts (schemas), not all of which are particularly easy to use or interpret (e.g. hierarchies are often but not always specified in `taxon_id,parent_id` pairs.)  To make it faster and easier to work across these providers, `taxadb` defines a common set of table schemas outlined below that are particularly suited for efficient computation of common tasks.  `taxadb` pre-processes and publicly archives compressed, flat tables corresponding to each of these schema for each of these providers.

Using the schema defined below, most common operations can be expressed in terms of very standard operations, such as simple filtering joins in SQL.  to implement these, `taxadb` imports the compressed flat files into a local, column-oriented database, [`MonetDBLite`](), which can be installed entirely as an R package with no additional server setup required.  This provides a persistent store, and ensures that operations can be performed on disk since the taxonomic tables considered here are frequently too large to store in active memory.  The columnar structure enables blazingly fast joins.  Once the database is created, `taxadb` simply wraps a set of user-friendly R functions around common `SQL` queries, implemented in the popular `dplyr` syntax.  By default, `taxadb` will always collect the results of these queries to return familiar, in-memory objects to the R user.  Optional arguments allow more direct access the database queries.  

This vignette summarizes the table schema defined by `taxadb`.  Pre-processing of the original database dumps from each provider into the format described here can be found in the corresponding scripts in `data-raw` directory of the R package source code. 


```{r message = FALSE}
library(dplyr)
library(taxadb)
td_create("all")
```

## data schema


- `taxonID`: a unique id for the name (including provider prefix).  Note that some providers do not assign IDs to synonyms, but only to accepted names.  In this case, the `taxonID` should be `NA`, and the ID to the accepted name should be specified in `acceptedNameUsageID`.  
- `scientificName`: a Latin name, either accepted or known synonym, at the lowest resolved level for the taxon.  While DWC encourages the use of authorship citations, these are intentionally omitted in most tables as inconsistency in abbreviations and formatting make names with authors much harder to resolve.  When available, this information is provided in the additional optional columns using the corresponding Darwin Core terms.  
- `taxonRank`: the rank (as given by the provider) of this taxon.
- `acceptedNameUsageID` the accepted identifier.  For synonyms, the scientificName of the row with the corresponding `taxonID` gives the accepted name, according to the data provider in question.  For accepted names, this is identical to the `taxonID` for the name. If not provided, it is assumed this is the same as the `taxonID`.  
- `taxonomicStatus` Either "accepted", for an accepted scientific name, or a term indicating if the name is a known synonym, common misspelling, etc.

Some providers may report additional optional columns, see below.  

```{r}
taxa_tbl("itis")
```


## Hierarchy Terms

Darwin Core defines several commonly recognized ranks as possible Taxon properties as well: `kingdom`, `phylum`, `class`, `order`, `family`, `genus`, `specificEpithet`, and `intraspecificEpithet`.  Additionally, the taxonomic rank of any scientific name can be specified under `taxonRank`, whether or not it is one of these names.  

Semantically (specifically in the RDF sense), treating ranks as properties seems somewhat crude.  Database providers (and thus different experts) disagree both about what rank levels they recognize and what names belong in what ranks.  NCBI recognizes over 40 named ranks and numerous unnamed ranks.  OTT, in true cladistic fashion, identifies all mammals as being not only in the class "Mammalia", but also in the "class" of lobe-finned-fish, Sarcopterygii.  To distinguish between these different treatments, it would be semantically most consistent to associate a (or multiple) `taxonRankID` with each taxonomic entry, rather than a a taxonRank. This ID could be specific to the data provider, and indicate the rank name that provider associates with that rank.  Few (wikidata, with its strong RDF roots, is an exception) providers associate IDs with rank levels though. 

In practice, treating ranks as properties (i.e. as column headings) is far more consistent with typical scientific usage and convenient for common applications, such as generating a list of all birds or all frogs by a simple filter on names in a column. 


## Synonyms

The `taxonomicStatus` value indicates if the name provided is a synonym, misspelling or an accepted name.  `taxadb` does not enforce any controlled vocabulary on the use of these terms beyond using the term `accepted` to indicate that the `scientificName` is an accepted name (i.e. the `dwc:acceptedNameUsage`) for the taxon.  Including both accepted names and synonyms in the `scientificName` column greatly facilitates taxonomic name resolution: a user can just perform an SQL filtering join from a given list of names and the taxadb table in order to resolve names to identifiers (`acceptedNameUsageID`s).  

Hierarchical information is given using the accepted name classification for synonyms as well as accepted names.  This makes it possible to filter a table by some major group:

```{r}
birds <- taxa_tbl() %>% filter(class=="Aves") 
birds %>% head()
```

before joining against either synonym or accepted name, and also makes it easier to see what the accepted name for a synonym:

```{r}
birds %>% 
  filter(taxonomicStatus != "accepted") %>% 
  select(scientificName, taxonomicStatus, genus, specificEpithet) %>%
  head()

```

## Additional fields


## Common names

Common names are available from several providers, but tidy tables for `taxadb` have not yet been implemented.  Common names tables are expected to follow the following schema:

- `id` The taxonomic identifier for the species (or possibly other rank)
- `name` The common name / vernacular name
- `language` The language in which the common name is given, if known. (all lowercase)
- `language_code` the two-letter language code.

## Linked Data formats

`taxadb` tables can easily be interpreted as semantic data and will be made available as RDF triples.  This permits the richer SPARQL-based queries of taxonomic information, in addition to the SQL-based queries.  This data format will be the focus of a separate R package interface `taxald`.  

## Conventions

- Identifiers use the integer identifier defined by the provider, prefixed by the provider abbreviation in all capital letters: `ITIS:`, `GBIF:`, etc.
- Rank names are always lower case without hyphens or spaces. Rank names should be mapped
  to a table of standard accepted rank names (i.e. those recognized by ITIS, NCBI, Wikidata),
  and rank names should have 
- Encoding is UTF-8



```{r include=FALSE}
taxadb:::td_disconnect()
```
